---
title: "Group_Report4_ModelFinal"
output: html_document
date: "2025-12-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# *Group work

# **Step 4 — Model Reflection & Interpretation**

# **Step 5 — Group Comparison & Synthesis (Collaborative)**

| Member | Adj R\^2 (train) | R\^2 (test) | RMSE | transformation | interactions/high-order | comment |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
|  |  |  |  |  |  |  |
|  |  |  |  |  |  |  |
|  |  |  |  |  |  |  |
|  |  |  |  |  |  |  |

# **Step 6 — Group Reflection (Collaborative)**

Throughout this project, we gained practical experience on the process of linear regression. We learned the importance of conducting a thorough EDA. Understanding the distributions and bivariate trends of the raw data helps make the process of model building go smoother. We also learned that transformations are only a possible solution to fixing residual analysis, not a guarantee. Because our raw data contained so much heteroscedasticity, the many combinations of transformations and higher order terms were not enough to make our residuals normal. This project showed us that it is important to be as in depth as possible in our EDA. If we would have noticed the heteroscedasticity in our raw data during our EDA, we could've spent more time looking for a different data set or exploring a different response variable for our analysis. If we had more time, we could have looked for and explored a different data set, now that we know our data set contains too much heteroscedasticity to be used for least-squares linear regression. Approaching this project by combining our different splits of data allowed us to test whether our individual findings were accurate. Across all members, our results were largely similar for all steps. There were slight differences in variable significance, adjusted \$R\^2\$ and RMSE, but overall, our models generalized moderately well while failing residual analysis. The collaborative synthesis was most helpful for residual analysis. All members having residuals that were not normal and did not have heteroscedasticity made it possible to conclude that our data didn't work well for linear regression rather than user error.

# **Step 7 — Final Submission Package (Collaborative)**

# *Individual work

# Kyle's work

```{r}
options(scipen = 7)

library(tidyverse)
library(ggplot2)
library(gridExtra)
library(polycor)
library(knitr)

residAnalysis <- function(model) {
  par(mfrow=c(2,2))
  plot(model)
  par(mfrow=c(1,1))
  
  cat("\nShapiro-Wilk's normality test for our model:\n")
  print(shapiro.test(residuals(model)))
  cat("\nDurbin-Watson's independence test for our model:\n\n")
  print(car::durbinWatsonTest(model))
}

modelAnalysis <- function(model, test_data){
  #Table for Train r^2, Test R^2, and RMSE
  
  train_r_2 <- summary(model)$adj.r.squared
  
  #Compute RMSE
  pred_test <- predict(model, newdata = test_data)
  RMSE <- sqrt(mean((test_data$expenditure - pred_test)^2))
  
  #Compute Test R^2
  ss_res <- sum((test_data$expenditure - pred_test)^2)
  ss_tot <- sum((test_data$expenditure - mean(test_data$expenditure))^2)
  test_r2 <- 1 - ss_res/ss_tot
  test_r2
  
  #Create Table
  model_values <- c(train_r_2, test_r2, RMSE)
  
  sum_table <- data.frame(model_values, row.names = c("Training R^2", "Test R^2", "RMSE"))
  kable(sum_table)
}

##Load data
set.seed(92394)

data <- read_csv("credit_card_data.csv", show_col_types = FALSE)

data <- data %>% filter(expenditure > 0)
n <- nrow(data)

train_index <- sample(1:n, size = 0.7 * n)
train_data <- data[train_index, ]
test_data  <- data[-train_index, ]
```


## Simplifying our model

Income and share provide reciprocal information, so we do not need to include both of them. We will remove income, since share has a higher correlation. A partial F-test would state that we should keep income, but that is because the reciprocal information is causing the linear regression to produce invalid results. Removing income should make our residuals act more normal and produce a model that is more accurate to the true regression line, even if it reduces our adjusted $R^2$ and RMSE.

```{r}
model1 <- lm(expenditure ~ sqrt(share) * dependents + months + active, data = train_data)
summary(model1)
modelAnalysis(model1, test_data)
```

Now we will see if months is important to keep in our model, since it is showing as insignificant in our model summary output.

```{r}
model2 <- lm(expenditure ~ sqrt(share) * dependents + active, data = train_data)
anova(model1, model2)
```

The parital F-test comparing the full model, model1, vs the reduced model, model2, reported a very high p-value of .9732. This means the difference in adding months as a predictor is not significant at all. We should exclude months and use the simpler model, model2.

```{r}
summary(model2)
modelAnalysis(model2, test_data)
```

We can see in the model summary output that our adjusted $R^2$ actually went up slightly by excluding months. This confirms that months was just adding random noise to our model.

## Residual and Assumption Review

```{r, echo = FALSE}
residAnalysis(model2)
```

Removing income and squaring share brought our residuals much closer to passing all assumptions. We can still see a slight pattern in our Residuals vs Fitted plot. Our Q-Q plot looks much better, besides deviations in the upper tail. Since we have already tried transformations on both expenditure and our response variables, our next option is to remove any outliers. We can do this for our data set because we want to predict expenditure for an average person using this credit card, not predict people who will spend more than the average person.

To remove outliers, we will look at standarized residuals and remove any that are over 3 standard deviations from 0. From our Q-Q plot, we can see 3 highlighted observations that are over 5 standard deviations from 0. We will start be removing these and checking our model again.

```{r}
train_data <- train_data[-c(342, 322, 492),]
model3 <- lm(expenditure ~ sqrt(share) * dependents + active, data = train_data)
summary(model3)
modelAnalysis(model3, test_data)
```

Our model performs slightly better now, with a training adjusted $R^2$ increasing to .7329, as well as the RMSE slightly reducing. Now we will see if the residual analysis improved as well.

```{r, echo = FALSE}
residAnalysis(model3)
```

The plots no longer have such extreme outliers, but the assumptions all still fail. This makes sense when looking back at our EDA. The bivariate relationships between expenditure and the predictor variables all contain heteroscedasticity. Creating a model without heteroscedasticity in the residuals is very unlikely when the data itself is full of it, especially if we want a model that is interpretative. 

## Prediction Accuracy

```{r}
modelAnalysis(model3, test_data)
```

This model performs better on the training data, with a slightly lower adjusted $R^2$ on the testing data. The RMSE, however, is still relatively low, showing that it is not completely useless in predicting expenditure. A test adjusted $R^2$ of .6774 and a RMSE of 139.306 shows that this model generalizes moderately well, with only slight instability between training and testing.

## Model Reflection and Interpretation

My final model is as follows:

$E[expenditure] = -204.6722 + 1491.9818(\sqrt{share}) + -10.0464(dependents) + 2.9297(active) + 146.9072(\sqrt{share})(dependents)$


The intercept is not useful, as a negative value for expenditure is not possible. Across all $B_1 \rightarrow B_4$, it is easy to see that $\sqrt{share}$ is the most important predictor variable. $B_1$ is much higher than all other predictors. The next highest is $B_4$, which is the interaction between $\sqrt{share}$ and dependents. Active was found to also have a small significant effect. Dependents on its own was not significant.

The most important change to our model was dropping income since it added redundant information and made the least squares method "break". Removing income made our residuals closer to normality and made our model and beta coefficients easier to interpret. When trying out different combinations of transformations, higher-order terms, and predictor variables, I obtained models that were slightly closer to passing residual analysis. None passed, however. Because none of my models passed residual analysis, I decided to used a much simpler model that was much more interpretable and still reasonably close to passing our residual analysis. Using the complex models would've left me with a model that did not satisfy the residual assumptions and would be useless practically, since the data was so transformed.


# Shreya's individual work
---
title: "Shreya Mohan Final Model"
author: "Shreya Mohan"
date: "2025-12-05"
output: html_document
---

### Data Import
```{r, echo = FALSE}
library(MASS)
library(lmtest)

credit_data <- read.csv("credit_card_data.csv")
head(credit_data)
set.seed(862336550)
n <- nrow(credit_data)
train_index <- sample(1:n, size = 0.7 * n)
train_data <- credit_data[train_index, ]
test_data  <- credit_data[-train_index, ]
```

### Week 8 Model:
```{r}
week8 <- lm(expenditure ~ income + sqrt(share) + dependents + months + active + share:dependents, data = train_data)
summary(week8)

res <- residuals(week8)
hist(res,
     main = "Histogram of Residuals",
     xlab = "Residuals",
     breaks = 30)
plot(week8, which = 1)
plot(week8, which = 2)
dwtest(week8)
```
The model I had last week did not satisfy the assumption for residuals or normality. Although a lot of the predictors were significant we wanted to run an anova comparison to see if we could simplify our model while still making accurate predictions. 

### Data Cleaning
```{r}
credit_data <- read.csv("credit_card_data.csv")
set.seed(862336550)
credit_data <- subset(credit_data, dependents != 0)
credit_data <- subset(credit_data, expenditure != 0)

n <- nrow(credit_data)

train_index <- sample(1:n, size = 0.7 * n)

train_data <- credit_data[train_index, ]

test_data  <- credit_data[-train_index, ]
nrow(credit_data)
nrow(train_data)
nrow(test_data)
```

Data cleaning was necessary as our data according to the data logs, dependents equals 1 + number of dependents. However, there is an error in the dataset in this case as we have observations where dependents = 0. Thus, we removed these observations for consistency. We also removed the observations where expenditure = 0 as these are meaningless for our predictions.

### Step 1 — Simplify Your Expanded Model (Individual Work)
#### Using the model you ended with last week (or your latest model) — which may include:

Here I removed income from last weeks model. This is because income and share provide reciprocal information. The partial F test said we should keeo income, but this is because  the reciprocal relationship is causing it to provide invalid results. We decided to keep share as it seemed more representative when predicting expenditure as it is the ratio of monthly credit card expenditure to yearly income. 

```{r}
model1 <- lm(expenditure ~ sqrt(share) + dependents + months + active + sqrt(share):dependents, data = train_data)
plot(model1, which = 1)
dwtest(model1)

model2 <- lm(expenditure ~ sqrt(share) + dependents + months + active, data = train_data)
plot(model2, which = 1)
dwtest(model2)

anova(model1, model2)
```

The partial F test shows a very small p value for the comparison between the full model (with interaction) and the reduced model (without interaction). This means that the interaction share:dependents is highly significant and that we should proceed with the full model. 

```{r}
model2 <- lm(expenditure ~ sqrt(share) + dependents + months + active + sqrt(share):dependents, data = train_data)

model3 <- lm(expenditure ~ sqrt(share) + dependents + active + sqrt(share):dependents, data = train_data)
dwtest(model3)

plot(model3, which = 1)

summary(model3)

anova(model2, model3)
```
The partial F-test comparing the full model (including months) to the reduced model (excluding months) yields a large p-value, indicating that months is not a significant predictor. Therefore, we can proceed with the reduced model without months.

```{r}
model3 <- lm(expenditure ~ sqrt(share) + dependents + active + sqrt(share):dependents, data = train_data)

model4 <- lm(expenditure ~ sqrt(share) + dependents + sqrt(share):dependents, data = train_data)
dwtest(model4)

plot(model4, which = 1)

summary(model4)

anova(model3, model4)
```
The partial F-test comparing the full model (including active) to the reduced model (excluding active) yields a large p-value, indicating that active is not a significant predictor. Therefore, we can proceed with the reduced model without active

```{r}
# Box-Cox transformation
boxcox(model4, lambda = seq(-2, 2, 0.1)) 
```

After running a boxcox on model4, we can see that $\lambda \approx 0.4$ which is close to 0.5 so we apply a square root transformation to our response variable expenditure. Below I applied that on our final model. I also removed the extreme outliers from the Normal Q-Q plot. 

```{r}
train_data <- train_data[-c(1189, 765, 295), ]
final_model <- lm(sqrt(expenditure) ~ sqrt(share) + dependents + sqrt(share):dependents, data = train_data)
summary(final_model)
```
### Step 2 — Residual & Assumption Review
```{r}
res <- residuals(final_model)
hist(res,
     main = "Histogram of Residuals",
     xlab = "Residuals",
     breaks = 30)

plot(final_model, which = 1)
plot(final_model, which = 2)
dwtest(final_model)
```
Applying the square root transformation improved the residual vs. fitted plot and the Q-Q plot, reducing the appearance of heteroskedasticity and making the model diagnostics more reasonable, even if not perfect. 

This improvement is consistent with our EDA, which showed that the relationships between expenditure and the predictor variables exhibit heteroskedasticity. Given the variability in the data, it is unrealistic to expect a model with perfectly homoskedastic residuals while still maintaining interpretability.

### Step 3 — Prediction Accuracy
```{r}
# Mean and SD of expenditure
mean <- mean(credit_data$expenditure)
sd   <- sd(credit_data$expenditure)

cat("Mean of Expenditure:", mean, "\n")
cat("SD of Expenditure:", sd, "\n\n")

# Predictions on test set (sqrt scale)
pred_test <- predict(final_model, newdata = test_data)

# Test RMSE
RMSE <- sqrt(mean((sqrt(test_data$expenditure) - pred_test)^2))
cat("Test RMSE:", RMSE, "\n\n")

# Training R^2 on sqrt scale
train_r2 <- summary(final_model)$r.squared
cat("Training R^2 (on sqrt scale):", train_r2, "\n")

ss_res <- sum((sqrt(test_data$expenditure) - pred_test)^2)
ss_tot <- sum((sqrt(test_data$expenditure) - mean(sqrt(test_data$expenditure)))^2)

test_r2 <- 1 - ss_res/ss_tot
cat("Testing R^2 (on sqrt scale):", test_r2, "\n")
```
The model generalizes reasonably well, with a test $R^2$ of 0.72, showing moderate predictive stability on new data. Though the high variability in expenditure, mean and standard deviation, the test RMSE of 4.34 indicates predictions are fairly close to observed values. The training $R^2$  of 0.84 and testing $R^2$  of 0.72 suggest that the model captures most of the variation in the data while maintaining reasonable performance on unseen observations.

### Step 4 — Model Reflection & Interpretation

The regression model predicting the square root of expenditure identified sqrt(share) and the interaction between sqrt(share) and dependents as the most important predictors. Both of these terms are highly significant, indicating that expenditure increases with share and that this effect is stronger for households with more dependents. The predictor dependents alone was not significant. Overall, the model explains a large portion of the variance with the training $R^2 = 0.84$ and testing $R^2 = 0.72$ and produces residuals that are roughly symmetric and unbiased, supporting the model assumptions.

Modeling decisions that influenced performance included applying a square root transformation to the response variable to stabilize variance and including the interaction between sqrt(share) and dependents, which captured a meaningful relationship not reflected by the main effects alone. Non-informative predictors such as months and active were removed based on partial F-tests, improving model interpretability without sacrificing the fit. This approach balanced predictive accuracy with simplicity, resulting in a model that is both statistically robust and easy to interpret for understanding how share and household dependents drive expenditure.

# Jenny's individual work

```{r}

library(car)
library(MASS)       

creditData <- read.csv("credit_card_data.csv")
set.seed(862389455)
rows <- nrow(creditData)
trainNum <- sample(1:rows, size = 0.7 * rows)
trainData <- creditData[trainNum, ]
testData  <- creditData[-trainNum, ]
```

```{r, include=FALSE}

# center points
  trainData$sqrt_share     <- sqrt(trainData$share)
  trainData$sqrt_months    <- sqrt(trainData$months)
  trainData$log_months    <- log(trainData$months + 1)
  trainData$log_income    <- log(trainData$income + 1)
  trainData$log_active   <- log(trainData$active + 1)
  trainData$log_dependents   <- log(trainData$dependents + 1)
  trainData$log_share   <- log(trainData$share + 1)

  
  trainData$sqrt_share_c   <- trainData$sqrt_share - mean(trainData$sqrt_share)
  trainData$sqrt_months_c  <- trainData$sqrt_months - mean(trainData$sqrt_months)
  trainData$log_months_c    <- trainData$log_months - mean(trainData$log_months)
  trainData$log_income_c    <- trainData$log_income - mean(trainData$log_income)
  trainData$log_active_c    <- trainData$log_active - mean(trainData$log_active)
  trainData$log_dependents_c    <- trainData$log_dependents - mean(trainData$log_dependents)
  trainData$log_share_c    <- trainData$log_share - mean(trainData$log_share)

  trainData$income_c     <- trainData$income - mean(trainData$income, na.rm = TRUE)
  trainData$share_c      <- trainData$share - mean(trainData$share, na.rm = TRUE)
  trainData$months_c     <- trainData$months - mean(trainData$months, na.rm = TRUE)
  trainData$dependents_c <- trainData$dependents - mean(trainData$dependents, na.rm = TRUE)
  trainData$active_c     <- trainData$active - mean(trainData$active, na.rm = TRUE)
                                              

mean_income      <- mean(trainData$income)
mean_share       <- mean(trainData$share)
mean_months      <- mean(trainData$months)
mean_active      <- mean(trainData$active)
mean_dependents  <- mean(trainData$dependents)

testData$log_income     <- log(testData$income + 1)
testData$log_dependents <- log(testData$dependents + 1)
testData$log_active     <- log(testData$active + 1)
testData$log_share      <- log(testData$share + 1)

testData$sqrt_share     <- sqrt(testData$share)
testData$sqrt_months    <- sqrt(testData$months)
  
testData$log_income_c      <- testData$log_income - mean(testData$log_dependents)
testData$log_dependents_c  <- testData$log_dependents - mean(testData$log_income)

testData$income_c          <- testData$income - mean_income
testData$share_c           <- testData$share - mean_share
testData$months_c          <- testData$months - mean_months
testData$active_c          <- testData$active - mean_active
testData$dependents_c      <- testData$dependents - mean_dependents
testData$log_share_c     <- testData$log_share - mean(testData$log_share)
testData$log_active_c     <- testData$log_active - mean(testData$log_active)
```

## Week 9

## 1.

Recap: We determined that the best model was the model from Week 8's step 4 with squared income, and an interaction variable between variables dependents and active.

```{r}
model2_poly_int_simplify <- lm(
  expenditure ~ income_c + share_c + I(income_c^2) + active_c + dependents_c + dependents_c * active_c, data = trainData
)
summary(model2_poly_int_simplify)
```

Now we can check using F tests if any variables can be removed to simplify the model.

```{r}
model_no_quad <- lm(
  expenditure ~ income_c + share_c + active_c +
    dependents_c + dependents_c * active_c,
  data = trainData
)

anova(model_no_quad, model2_poly_int_simplify)

```

The partial F-test comparing the full model to a reduced model without the quadratic income term showed that removing income squared did not significantly reduce model fit as F(1,916)=0.88,p=0.349. Because the quadratic component of income does not meaningfully improve prediction and is not statistically supported, we can remove the quadratic term for parsimony.

To further simplify our model we can proceed to remove income as it is very similar to share, and out of the two share is slightly more informative and may be able to provide us with more insights into our model.

```{r}
model_no_quad <- lm(
  expenditure ~ income_c+ share_c + active_c + dependents_c + dependents_c * active_c, data = trainData
)

summary(model_no_quad)

```

```{r}
  pred_old <- predict(model_no_quad, newdata = trainData)
  rmse_old <- sqrt(mean((trainData$expenditure - pred_old)^2, na.rm = TRUE))
  rmse_old
  
  pred_old <- predict(model2_poly_int_simplify, newdata = trainData)
  rmse_old <- sqrt(mean((trainData$expenditure - pred_old)^2, na.rm = TRUE))
  rmse_old
```

Removing our quadratic term income squared did not change the performance of our model all that much, so taking it out manages to simplify the model while maintaining generally the same reliability.

## 2.

Now we proceed to check our residuals vs fitted model, QQ plot, etc.

```{r}
  model_final <- model_no_quad
  
  plot(model_final, which = 1, col = "orange")
  
  
  plot(model_final, which = 2, col = "orange")
  
# checking if outliers shuld be removed
  plot(model_final, which = 4, col = "orange")   
  plot(model_final, which = 5, col = "orange")  
  plot(model_final, which = 6, col = "orange") 
```

Our residuals vs fitted and QQ residuals plots have stayed relatively the same compared to our simplified model, which is a positive note. However, looking at our Cook's distance and Residuals vs Leverage plots, we can see that there are some points that may be described as highly influential. This should be looked into and removal of outliers should be considered.

```{r}

std_res <- rstandard(model_final)
outliers <- order(std_res, decreasing = TRUE)[1:3]
outliers

trainData_no_out <- trainData[-outliers, ]


# Refit the model
model_final_no_out <- lm(
  expenditure ~ income_c + share_c + active_c + dependents_c + dependents_c * active_c,
  data = trainData_no_out
)


summary(model_final_no_out)


# Residual vs fitted plot
par(mar = c(4,4,2,1))
plot(model_final_no_out, which = 2, col = "orange")
plot(model_final_no_out, which = 1, col = "orange")

 
  pred_out <- predict(model_final_no_out, newdata = trainData_no_out)
  rmse_out <- sqrt(mean((trainData_no_out$expenditure - pred_out)^2, na.rm = TRUE))
  rmse_out
```

Our RMSE and residuals have stayed relatively the same, but our adjusted R\^2 has showed very minimal, almost insignificant improvement.

## 3. 
```{r}
pred_test <- predict(model_final_no_out, newdata = testData)
RMSE <- sqrt(mean((testData$expenditure - pred_test)^2))
RMSE
summary(model_final_no_out)$r.squared




```

The RMSE of our model has gone up by around 50%, but our adjusted R\^2 has stayed relatively the same, maintaining a generally reliable accuracy.

## 4.

The final regression model, expenditure \~ income_c + share_c + active_c + dependents_c + dependents_c \* active_c identifies income, insurance share, and the dependents × active interaction as the strongest predictors. Both income and insurance share show large, positive, and statistically significant effects. The interaction between dependents and active status remains significant, suggesting that the influence of dependents on expenditure depends on whether an individual is active.

Modeling decisions that had the greatest influence included removing unneeded nonlinear terms, which still managed to keep the meaningful interactions. After removing a small number of extreme outliers identified in the Q–Q plot, predictive accuracy decreased (RMSE increased from 102.22 to 150.25), but overall fit remained effectively unchanged. This shows the balance between fit and interpretability as we wanted to keep a simple, interpretable model while keeping crucial components. Despite the increase in RMSE, the cleaned model can still provide a relatviely reliable representation

# Lydia's individual work
