---
title: "Shreya Mohan Final Model"
author: "Shreya Mohan"
date: "2025-12-05"
output: html_document
---

### Data Import
```{r, echo = FALSE}
library(MASS)
library(lmtest)

credit_data <- read.csv("credit_card_data.csv")
head(credit_data)
set.seed(862336550)
n <- nrow(credit_data)
train_index <- sample(1:n, size = 0.7 * n)
train_data <- credit_data[train_index, ]
test_data  <- credit_data[-train_index, ]
```

### Week 8 Model:
```{r}
week8 <- lm(expenditure ~ income + sqrt(share) + dependents + months + active + share:dependents, data = train_data)
summary(week8)

res <- residuals(week8)
hist(res,
     main = "Histogram of Residuals",
     xlab = "Residuals",
     breaks = 30)
plot(week8, which = 1)
plot(week8, which = 2)
dwtest(week8)
```
The model I had last week did not satisfy the assumption for residuals or normality. Although a lot of the predictors were significant we wanted to run an anova comparison to see if we could simplify our model while still making accurate predictions. 

### Data Cleaning
```{r}
credit_data <- read.csv("credit_card_data.csv")
set.seed(862336550)
credit_data <- subset(credit_data, dependents != 0)
credit_data <- subset(credit_data, expenditure != 0)

n <- nrow(credit_data)

train_index <- sample(1:n, size = 0.7 * n)

train_data <- credit_data[train_index, ]

test_data  <- credit_data[-train_index, ]
nrow(credit_data)
nrow(train_data)
nrow(test_data)
```

Data cleaning was necessary as our data according to the data logs, dependents equals 1 + number of dependents. However, there is an error in the dataset in this case as we have observations where dependents = 0. Thus, we removed these observations for consistency. We also removed the observations where expenditure = 0 as these are meaningless for our predictions.

### Step 1 — Simplify Your Expanded Model (Individual Work)
#### Using the model you ended with last week (or your latest model) — which may include:

Here I removed income from last weeks model. This is because income and share provide reciprocal information. The partial F test said we should keeo income, but this is because  the reciprocal relationship is causing it to provide invalid results. We decided to keep share as it seemed more representative when predicting expenditure as it is the ratio of monthly credit card expenditure to yearly income. 

```{r}
model1 <- lm(expenditure ~ sqrt(share) + dependents + months + active + sqrt(share):dependents, data = train_data)
plot(model1, which = 1)
dwtest(model1)

model2 <- lm(expenditure ~ sqrt(share) + dependents + months + active, data = train_data)
plot(model2, which = 1)
dwtest(model2)

anova(model1, model2)
```

The partial F test shows a very small p value for the comparison between the full model (with interaction) and the reduced model (without interaction). This means that the interaction share:dependents is highly significant and that we should proceed with the full model. 

```{r}
model2 <- lm(expenditure ~ sqrt(share) + dependents + months + active + sqrt(share):dependents, data = train_data)

model3 <- lm(expenditure ~ sqrt(share) + dependents + active + sqrt(share):dependents, data = train_data)
dwtest(model3)

plot(model3, which = 1)

summary(model3)

anova(model2, model3)
```
The partial F-test comparing the full model (including months) to the reduced model (excluding months) yields a large p-value, indicating that months is not a significant predictor. Therefore, we can proceed with the reduced model without months.

```{r}
model3 <- lm(expenditure ~ sqrt(share) + dependents + active + sqrt(share):dependents, data = train_data)

model4 <- lm(expenditure ~ sqrt(share) + dependents + sqrt(share):dependents, data = train_data)
dwtest(model4)

plot(model4, which = 1)

summary(model4)

anova(model3, model4)
```
The partial F-test comparing the full model (including active) to the reduced model (excluding active) yields a large p-value, indicating that active is not a significant predictor. Therefore, we can proceed with the reduced model without active

```{r}
# Box-Cox transformation
boxcox(model4, lambda = seq(-2, 2, 0.1)) 
```

After running a boxcox on model4, we can see that $\lambda \approx 0.4$ which is close to 0.5 so we apply a square root transformation to our response variable expenditure. Below I applied that on our final model. I also removed the extreme outliers from the Normal Q-Q plot. 

```{r}
train_data <- train_data[-c(1189, 765, 295), ]
final_model <- lm(sqrt(expenditure) ~ sqrt(share) + dependents + sqrt(share):dependents, data = train_data)
summary(final_model)
```
### Step 2 — Residual & Assumption Review
```{r}
res <- residuals(final_model)
hist(res,
     main = "Histogram of Residuals",
     xlab = "Residuals",
     breaks = 30)

plot(final_model, which = 1)
plot(final_model, which = 2)
dwtest(final_model)
```
Applying the square root transformation improved the residual vs. fitted plot and the Q-Q plot, reducing the appearance of heteroskedasticity and making the model diagnostics more reasonable, even if not perfect. 

This improvement is consistent with our EDA, which showed that the relationships between expenditure and the predictor variables exhibit heteroskedasticity. Given the variability in the data, it is unrealistic to expect a model with perfectly homoskedastic residuals while still maintaining interpretability.

### Step 3 — Prediction Accuracy
```{r}
# Mean and SD of expenditure
mean <- mean(credit_data$expenditure)
sd   <- sd(credit_data$expenditure)

cat("Mean of Expenditure:", mean, "\n")
cat("SD of Expenditure:", sd, "\n\n")

# Predictions on test set (sqrt scale)
pred_test <- predict(final_model, newdata = test_data)

# Test RMSE
RMSE <- sqrt(mean((sqrt(test_data$expenditure) - pred_test)^2))
cat("Test RMSE:", RMSE, "\n\n")

# Training R^2 on sqrt scale
train_r2 <- summary(final_model)$r.squared
cat("Training R^2 (on sqrt scale):", train_r2, "\n")

ss_res <- sum((sqrt(test_data$expenditure) - pred_test)^2)
ss_tot <- sum((sqrt(test_data$expenditure) - mean(sqrt(test_data$expenditure)))^2)

test_r2 <- 1 - ss_res/ss_tot
cat("Testing R^2 (on sqrt scale):", test_r2, "\n")
```
The model generalizes reasonably well, with a test $R^2$ of 0.72, showing moderate predictive stability on new data. Though the high variability in expenditure, mean and standard deviation, the test RMSE of 4.34 indicates predictions are fairly close to observed values. The training $R^2$  of 0.84 and testing $R^2$  of 0.72 suggest that the model captures most of the variation in the data while maintaining reasonable performance on unseen observations.

### Step 4 — Model Reflection & Interpretation

The regression model predicting the square root of expenditure identified sqrt(share) and the interaction between sqrt(share) and dependents as the most important predictors. Both of these terms are highly significant, indicating that expenditure increases with share and that this effect is stronger for households with more dependents. The predictor dependents alone was not significant. Overall, the model explains a large portion of the variance with the training $R^2 = 0.84$ and testing $R^2 = 0.72$ and produces residuals that are roughly symmetric and unbiased, supporting the model assumptions.

Modeling decisions that influenced performance included applying a square root transformation to the response variable to stabilize variance and including the interaction between sqrt(share) and dependents, which captured a meaningful relationship not reflected by the main effects alone. Non-informative predictors such as months and active were removed based on partial F-tests, improving model interpretability without sacrificing the fit. This approach balanced predictive accuracy with simplicity, resulting in a model that is both statistically robust and easy to interpret for understanding how share and household dependents drive expenditure.
